Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2018-06-12 articles/formalism/sections/real_space_formalism.tex
Edits and queries in purifying section

We can only measure the total Stokes vector, a sum of the part corresponding to scalar $E$ and the part corresponding to $B$.  The $E$/$B$ modes are orthogonal to each other in the sense that their respective operators are orthogonal to each other as discussed in \sec{sec:mat_pol_intro}. It is possible to decompose the Stokes vector \vp{} into one \vp{\rm E} that purely contributes to $E$ modes and another \vp{\rm B} that purely contribute to the $B$ modes of polarization. In this section we derive the real space operators which operate on the total Stokes vector and yield this decomposition, without ever having to explicitly evaluate the scalar modes. The algebra is more involved, but the derivation is similar to that discussed in \sec{sec:qu2eb}, so we refrain from presenting the detailed calculations here, and outline only the key points. We use the harmonic space projection operators $\tilde O_{E/B}$, defined in \eq{eq:har_eb_op}, to derive the respective real space operators. The Stokes parameters corresponding to each scalar mode are given by the following expressions:

@@

where $\mathcal{I}_{r} ~\&~ \mathcal{D}_{r}$ and $\mathcal{I}_{i} ~\&~ \mathcal{D}_{i}$ are the real and complex parts of the following complex functions:

@@

where all the symbols have their usual meaning and the explicit multipole dependence of the real space operators has been suppressed for brevity. \revisit{Note that when the operators are expressed in terms of the Euler angles $(\alpha_{qe},\beta_{qe},\gamma_{qe})$ they can be interpreted as the Greens functions and  we denote them by $\mathcal{I}_G=\Delta \Omega \mathcal{I}^*$ and $\mathcal{D}_G=\Delta \Omega \mathcal{D}$. When expressed as function of Euler angles $(\alpha_{eq},\beta_{eq},\gamma_{eq})$ corresponding to the inverse rotations they can be interpreted as some convolving beam and we denote them by $\mathcal{I}_B=\Delta \Omega \mathcal{I}$ and $\mathcal{D}_B=\Delta \Omega \mathcal{D}$.} Note that unlike in the case of the operators $\mm_G$ and $\mm_B$ which have different shapes owing to their dependence on Euler angles $\alpha$ and $\gamma$ respectively, the operators $D_G$ and $D_B$ are identical since $(\alpha_{qe}-\gamma_{qe}) = (\alpha_{eq}-\gamma_{eq})$, while $\mathcal{I}_{G}$ and $\mathcal{I}_B$ are related by conjugation since  $(\alpha_{qe}+\gamma_{qe}) = -(\alpha_{eq}+\gamma_{eq})$.

The operator $\mathcal{I}$ is Hermitian and is a band limited version of the delta function owing to the identity: $\lim_{\ell \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$. For all practical purposes $\mathcal{I}$ acts like an identity operator as ascertained by the following set of identities: (i) $\mathcal{I} \mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D} \mathcal{I}=\mathcal{D}$. $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{D}^*$ is its inverse in this band limited sense: $\mathcal{D}^* \mathcal{D}=\mathcal{I}$. Using these properties\footnote{While testing the real space operator identities one encounters terms like $\mathcal{D} \mathcal{I}^*,\mathcal{I}^*\mathcal{I}$ and $\mathcal{I}\mathcal{I}^*$ which cannot be simply interpreted but they always occur in pairs with opposite signs that exactly cancel each other.} of the operators $\mathcal{I}$ and $\mathcal{D}$ , one can verify that the real space operators satisfy the following identities:
%
\begin{subequations}
\beqry
\bar O_E \, \bar O_E &=& \bar O_E \qquad ;\qquad \bar O_B \, \bar O_B = \bar O_B \,, \\
\bar O_E \, \bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of their harmonic space counterparts discussed in \sec{sec:mat_pol_intro}. Thus they are exactly orthogonal and idempotent. Note that unlike in the harmonic case, the sum of the operators is the band limited identity operator $\mathcal{I}$. This non-exactness is representative of the loss of information resulting from making this transformation on measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators which is exact and a more crucial property of the operators.

>>>>

We can only measure the total Stokes vector, a sum of the part that corresponds to scalar $E$ and the part that corresponds to $B$.  The $E$/$B$ modes are orthogonal to each other in the sense that their respective operators are orthogonal to each other as discussed in \sec{sec:mat_pol_intro}. It is possible to decompose the Stokes vector \vp{} into one \vp{\rm E} that purely contributes to $E$ modes and another \vp{\rm B} that purely contribute to the $B$ modes of polarization. In this section we derive the real space operators which operate on the total Stokes vector and yield this decomposition, without ever having to explicitly evaluate the scalar modes. The algebra is more involved, but the derivation is similar to that discussed in \sec{sec:qu2eb}, so we refrain from presenting the detailed calculations here, and outline only the key points. We use the harmonic space projection operators $\tilde O_{E/B}$, defined in \eq{eq:har_eb_op}, to derive the respective real space operators. The Stokes parameters corresponding to each scalar mode are given by the following expressions:

@@

where $\mathcal{I}_{r}, \mathcal{D}_{r}$ and $\mathcal{I}_{i}, \mathcal{D}_{i}$ are the real and imaginary parts of the following complex functions:

@@

where all the symbols have their usual meaning and the explicit multipole dependence of the real space operators has been suppressed for brevity. \revisit{Note that when the operators are expressed in terms of the Euler angles $(\alpha_{qe},\beta_{qe},\gamma_{qe})$ they can be interpreted as the Greens functions and  we denote them by $\mathcal{I}_G=\Delta \Omega \mathcal{I}^*$ and $\mathcal{D}_G=\Delta \Omega \mathcal{D}$. When expressed as function of Euler angles $(\alpha_{eq},\beta_{eq},\gamma_{eq})$ corresponding to the inverse rotations they can be interpreted as some convolving beam and we denote them by $\mathcal{I}_B=\Delta \Omega \mathcal{I}$ and $\mathcal{D}_B=\Delta \Omega \mathcal{D}$. [K: question about $\Delta \Omega$ here.]} Note that unlike in the case of the operators $\mm_G$ and $\mm_B$ which have different shapes owing to their dependence on Euler angles $\alpha$ and $\gamma$ respectively, the operators $D_G$ and $D_B$ are identical since $(\alpha_{qe}-\gamma_{qe}) = (\alpha_{eq}-\gamma_{eq})$, while $\mathcal{I}_{G}$ and $\mathcal{I}_B$ are related by conjugation since  $(\alpha_{qe}+\gamma_{qe}) = -(\alpha_{eq}+\gamma_{eq})$.

The operator $\mathcal{I}$ is Hermitian and is a band limited version of the delta function owing to the identity: $\lim_{\ell_{\rm max} \rightarrow \infty} \mathcal{I} = \delta(\hat{n}_i - \hat{n}_j)$. For all practical purposes $\mathcal{I}$ acts like an identity operator as ascertained by the following set of identities: (i) $\mathcal{I} \mathcal{I}=\mathcal{I}$ ; (ii) $\mathcal{D} \mathcal{I}=\mathcal{D}$. $\mathcal{D}$ is a complex but symmetric matrix and $\mathcal{D}^*$ is its inverse in this band limited sense: $\mathcal{D}^* \mathcal{D}=\mathcal{I}$. Using these properties\footnote{While testing the real space operator identities one encounters terms like $\mathcal{D} \mathcal{I}^*,\mathcal{I}^*\mathcal{I}$ and $\mathcal{I}\mathcal{I}^*$ which cannot be simply interpreted but they always occur in pairs with opposite signs that exactly cancel each other.} of the operators $\mathcal{I}$ and $\mathcal{D}$ , one can verify that the real space operators satisfy the following identities:
%
\begin{subequations}
\beqry
\bar O_E \, \bar O_E &=& \bar O_E;\qquad \bar O_B \, \bar O_B = \bar O_B \,, \\
\bar O_E \, \bar O_B &=& 0 \,,\\
\bar O_E + \bar O_B &=& \mathcal{I} \,,
\eeqry
\end{subequations}
%
which are the real space analogues of their harmonic space counterparts discussed in \sec{sec:mat_pol_intro}. Thus they are exactly orthogonal and idempotent. Note that unlike in the harmonic case, the sum of the operators is the band limited identity operator $\mathcal{I}$. This non-exactness is representative of the loss of information resulting from making this transformation on measured data with some imposed band limit. Forcing the sum of the operators to be exactly an identity matrix compromises the orthogonality property of the $\bar{O}_E$ \& $\bar{O}_B$ operators which \revisit{is exact and a more crucial property of the operators.[K: what does this mean?]}
